---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

> Adapted from Federico Abascal's practical

# Detection of drivers in bladder carcinoma

We will be working with a dataset of bladder cancer from the TCGA
consortium.

## Generating the input ﬁle

dndscv works with an input ﬁle that consist of 5 columns. 1) Sample ID
2) Chromosome 3) Position 4) Reference 5) Mutation

We can extract this data from MAF ﬁles columns:

1.  Tumor_Sample_Barcode
2.  Chromosome
3.  Start_Position
4.  Reference_Allele
5.  Tumor_Seq_Allele2

The input ﬁle can also be generated from vcf ﬁles using the following
columns:

1.  Tumor column (column number can diﬀer based on the tool used to
    generate the vcf ﬁle)
2.  Chrom
3.  POS
4.  REF
5.  ALT

For this practical we have already prepared an input ﬁle for you
TCGA-BLCA.5col. However, in your own work, you will have to typically
generate this input ﬁle on your own. The easiest way to structure this
input is by converting your variant calling output (.vcf formatted ﬁle)
to .maf format. (See earlier discussion on the topic)

## Data loading and exploration

Once we have an appropriately formatted 5 column input ﬁle with a
suitable header we should load the data and explore it to ensure that
the data are what we expect them to be and to check for any unexpected
errors.

### Load the input ﬁle

```{r}
library(dndscv)

setwd("~/Downloads/Bioinformatica_Avanzada/mutational_signatures/BLCA-mutational-signatures-2")

muts = read.table("example_BLCA/TCGA-BLCA.5col", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(muts)
```

To count the number of samples we can run the next command

```{r}
length(unique(muts$sampleID))
```

and to get the number of mutations:

```{r}
nrow(muts)
```

There are $370$ donors and $53518$ mutations. To see the mutational
burden per sample we can run the following:

```{r}
barplot(sort(table(muts$sampleID)), ylab = "Number of mutations", xlab = "Donors",  las = 2, names.arg = "")
```

### Are there any hypermutators in the cohort?

> *It is relevant to explore our data in this way because hypermutators
> can have a negative impact on the statistical power to detect drivers
> and because some hypermutators (e.g. POLE mutant tumors) are under
> mutational processes not properly modeled by a
> trinucleotide-substitution model. Although there is no exact deﬁnition
> of a hypermutator, usually having more than 500 mutations in the exome
> (approximately 5.6810\^{4} mutations in the whole genome) can be
> considered a hypermutator. In general, it is good to exclude these
> samples from this analysis.*

```{r}
donor_muts = table(muts$sampleID)
(hypmutated = donor_muts[donor_muts > 500])
```

The donors TCGA-G2-A3VY, TCGA-G2-AA3B and TCGA-ZF-AA4V are hypermutated.

## Driver detection

We will run dndscv to detect driver genes in bladder cancer removing
hypermutators (n\>500)

```{r}
BLCA_dndscv = dndscv(muts, max_muts_per_gene_per_sample = 3, max_coding_muts_per_sample = 500, outmats = TRUE)
```

While running dndscv you will se some warnings. One of them reads "Same
mutations observed in diﬀerent sampleIDs. Please verify that these are
independent events and remove duplicates otherwise." This warning
relates to that only unique mutations are listed in the input ﬁle. For
example if your input contains the same mutation in several related
samples (samples that come from the same tumor) they should only be
listed once in the ﬁle. You will also see a warning indicating that some
mutations have a wrong reference. This is because of a error in the
original TCGA ﬁle. We can ignore this as the number of aﬀected bases is
very small

### Looking at the output

dndscv generates a list of objects as output. You can look at the
contents of the list like this:

```{r}
names(BLCA_dndscv)
```

The most relevant output often is **sel_cv** as it contains the results
of the neutrality tests at gene level. The **globaldnds** output has a
table with global MLEs for the dN/dS ratios across all genes and their
conﬁdence intervals. The **annotmuts** output contains a table with
annotated coding mutations. **genemuts** has a table with observed and
expected number of mutations per gene.

### Table of significant genes

**BLCA_dndscv\$sel_cv** contains the results for all the analyzed genes.

We can look at the significant genes in the table by filtering for genes
with a **qglobla_cv \< 0.1**. **qglobal** is the multiple hypothesis
correction q-value for the **pglobal_cv** (the combined p-value for the
different p-values calculated).

```{r}
(driver_genes_BLCA = BLCA_dndscv$sel_cv[which(BLCA_dndscv$sel_cv$qglobal_cv < 0.1),])
```

#### How many significant genes do you find?

```{r}
nrow(driver_genes_BLCA)
```

There are 33 significant genes.

#### Is there any gene under negative selection?

In this list there aren't. This is because in all genes at least one
type of mutation is present in a much higher rate than that expected
under neutrality. In negative selection we would be expecting to see
genes that can't be mutated for the sake of the cancer cell survival,
and thus the mutation rate should be close to 0. There aren't any genes
that follow these conditions.

#### Which genes are oncogenes? Which genes are tummor suppresors?

> *Tip: look at the number and types of mutations n_syn, n_mis, ...*

```{r}
sup_genes = c()
onc_genes = c()
  
for ( gene in 1:nrow(driver_genes_BLCA)) {
  
  sup_score = driver_genes_BLCA[gene,]$wnon_cv + driver_genes_BLCA[gene,]$wspl_cv 
  onc_score = driver_genes_BLCA[gene,]$wmis_cv
  
  if (sup_score > onc_score){
    sup_genes = c(sup_genes, driver_genes_BLCA[gene,]$gene_name)
  } else {
    onc_genes = c(onc_genes, driver_genes_BLCA[gene,]$gene_name)
  }
  
}

cat("These are tummor suppresors: ", sup_genes, "\n")
cat("These are oncogenes: ", onc_genes)
```

#### Considering the coefficient of selection for missense mutations in ARID1A how many missense mutations had been selected for this cohort?

> Tip: the coeﬃcient wmis_cvis 4.4825244 and there are 21 missense
> mutations in ARID1A. Tip 2: (w-1)/w gives the proportion under
> positive selection. Tip 3: 95% conﬁdence intervals for the selection
> coeﬃcients can be obtained with **geneci(dout,gene_list="ARID1A")**
> Tip 4: Have a look at genemuts to see how many mutations were
> expected\*

```{r}
BLCA_dndscv$genemuts[which(BLCA_dndscv$genemuts$gene_name=="ARID1A"),]
geneci(BLCA_dndscv, gene_list = "ARID1A")
```

```{r}
# The number of missense mutations used to get the value of wmis_cvis is the number of obersved missense mutations minus the expected number of missense mutations under neutrality.

(selec_miss = BLCA_dndscv$genemuts[which(BLCA_dndscv$genemuts$gene_name=="ARID1A"),]$n_mis - BLCA_dndscv$genemuts[which(BLCA_dndscv$genemuts$gene_name=="ARID1A"),]$exp_mis)
```

Around 15-16 missense mutations were used in this cohort to estimate its
coefficient of selection

#### Are all of those mutations under selection?

No, not all of them are under selection, because the model already
expected to have 5-6 missense mutations assuming neutrality towards
cancer.

Let's take a look at the mutations in *PIK3CA*:

```{r}
BLCA_dndscv$annotmuts[which(BLCA_dndscv$annotmuts$gene=="PIK3CA"),]
```

Look at the aachange column to see the amino acid changes generated by
the mutations.

#### Is there any recurrent mutation (hotspot)?

```{r}
table(BLCA_dndscv$annotmuts[which(BLCA_dndscv$annotmuts$gene=="PIK3CA"),]$aachange)
```

Yes, there are two recurring missense mutations or hotspots, E545K and
E542K.

### Global signs of selection

dndscv also estimates global dN/dS ratios in aggregate for all the
genes. You can access this data in the dndscv output.

```{r}
print(BLCA_dndscv$globaldnds)
```

wspl is lower than 1. That could mean negative selection but this result
is often obtained with exome data because of the poorer sequencing
coverage at splice sites. **dndscv** interprets the depletion of mutations at splice sites as negative selection.

However all the other coeﬃcients are \> 1 and their 95% conﬁdence
intervals too.

We can use the **globaldnds** information to estimate the number of
missense driver mutations per sample.

-   There are 30610 missense mutations in the cohor, and the coefficient
    of selection **wmis** is 1.0524248.
-   Calculate the proportion of missense mutations under positive selection using the formula **(w-1)/w**.
- Find out the actual number of missense mutations under positive selection $n_mis * (w-1)/w$.
- Calculate the average per sample: $( n_mis * (w-1)/w ) / num_samples$.
You can obtain all the info with:

```{r}
w = BLCA_dndscv$globaldnds[1,2]
n_mis = length(which(BLCA_dndscv$annotmuts$impact=="Missense"))
(num_samples = length(unique(muts$sampleID)))

#Proportion of missense mutations under positive selection
pos_sel_mis = (w - 1)/w
```

